\section{System Architecture}

The architectural evolution investigated in this work is illustrated in \autoref{fig:workflow_vergleich}, which contrasts two conceptually distinct system designs. Version 1 (V1) represents a naive RAG pipeline, while Version 2 (V2) reflects a transition toward a more advanced and partially modular architecture inspired by recent developments in RAG research.

As shown in \autoref{fig:workflow_vergleich}(a), V1 follows a strictly linear workflow consisting of contextualization, semantic retrieval, generation, and postprocessing. This design corresponds to what the literature describes as Naive RAG, where retrieved documents are directly appended to the prompt without further refinement or adaptive control. While such architectures are straightforward to implement, they exhibit well-documented limitations, particularly in retrieval precision, handling of complex queries, and robustness against irrelevant context. These limitations were also evident in the analyzed system, which relied on rigid workflows, limited ingestion capabilities, and hardcoded logic.

In contrast, \autoref{fig:workflow_vergleich}(b) depicts the redesigned architecture (V2), which introduces a routing-based control layer that dynamically orchestrates multiple processing paths. This shift reflects the broader movement from static pipelines toward adaptive RAG systems. The Router \& Contextualizer serves as the central decision-making component, classifying incoming queries and directing them into one of three paths: (1) a \textit{Path without Retrieval}, (2) an augmented \textit{Q\&A} path, and (3) a pedagogically motivated \textit{Socratic} learning mode.
The introduction of a no-retrieval path is directly motivated by recent findings on adaptive retrieval strategies. Simple or conversational queries can be answered without accessing external knowledge, reducing latency and computational cost while avoiding unnecessary retrieval noise. This replaces the previously hardcoded no-answer logic with a more flexible and context-aware mechanism .
For knowledge-intensive queries, the system employs a hybrid retrieval pipeline followed by a reranking stage, as shown in \autoref{fig:workflow_vergleich}(b). This design choice is strongly grounded in the literature, which consistently identifies retrieval quality as a primary determinant of system performance. Due to constraints in modifying underlying models, reranking is implemented via an additional LLM-based scoring step. Furthermore this workflow employs a query decomposition strategy for queries that require multi-hop reasoning. Notably, more complex architectural paradigms such as GraphRAG, iterative agent-based systems (e.g., SELF-RAG), or fully modular orchestration frameworks were deliberately not implemented. While these approaches offer advantages for multi-hop reasoning and global knowledge representation, they introduce significant computational overhead, increased system complexity, and higher maintenance requirements. Given the applied project context and the need for controlled comparative evaluation, a more constrained advanced RAG design was prioritized. Similarly, iterative retrieval loops and reinforcement-based feedback mechanisms were excluded due to their implementation complexity and unclear short-term benefit within the given resource constraints.
A distinctive extension of V2 is the integration of a Socratic learning assistant, represented as a separate processing branch in \autoref{fig:workflow_vergleich}(b). Unlike standard Q\&A interactions, this mode guides learners through structured questioning rather than direct answer delivery. The interaction follows staged phases including goal elicitation, diagnostic questioning, iterative reasoning, and feedback, enabling a shift from information retrieval toward active knowledge construction. This design operationalizes pedagogical insights from educational research while remaining compatible with the overall RAG architecture.

\begin{figure}[htbp]
\centering

\tikzset{
    box/.style={
        draw,
        rectangle,
        fill=white,
        text width=3.0cm,
        minimum height=1.15cm,
        align=center
    },
    line/.style={
        -Latex,
        thick
    }
}

% ===================== VERSION 1 =====================
\begin{minipage}{\textwidth}
\centering
\begin{tikzpicture}[
    font=\small,
    >=Latex
]

\node[box] (c1) at (1.5,0)  {Contextualizer};
\node[box] (r1) at (5.2,0)  {Semantic\\Retrieval};
\node[box] (g1) at (8.9,0)  {Generator};
\node[box] (p1) at (12.6,0) {Postprocessing};

\draw[line] (c1.east) -- (r1.west);
\draw[line] (r1.east) -- (g1.west);
\draw[line] (g1.east) -- (p1.west);

\end{tikzpicture}

\vspace{0.4em}
\textbf{(a) Version 1}
\end{minipage}

\vspace{1.2em}

% ===================== VERSION 2 =====================
\begin{minipage}{\textwidth}
\centering
\begin{tikzpicture}[
    font=\small,
    >=Latex
]

% Router
\node[box] (router) at (1.8,0) {Router \&\\Contextualizer};

% Split point
\coordinate (split) at (4.0,0);

% Upper workflow
\node[box] (gen1) at (14.2,1.8) {Generator};

% Middle workflow
\node[box] (retrieval) at (6.8,0)  {Hybrid\\Retrieval};
\node[box] (rerank)    at (10.5,0) {Reranker};
\node[box] (gen2)      at (14.2,0) {Generator \&\\Postprocessing};

% Lower workflow
\node[box] (socraticmode) at (8.6,-1.8) {Socratic learning\\mode};

% Trunk
\draw[thick] (router.east) -- (split);

% Branches from split
\draw[line] (split) |- (gen1.west);
\draw[line] (split) -- (retrieval.west);
\draw[line] (split) |- (socraticmode.west);

% Middle workflow
\draw[line] (retrieval.east) -- (rerank.west);
\draw[line] (rerank.east) -- (gen2.west);

% Labels at branching point
\node[anchor=west, align=left] at (4,2) {Path without Retrieval};
\node[anchor=west]             at (4,0.21) {Q\&A};
\node[anchor=west]             at (4,-1.57) {Socratic};

\end{tikzpicture}

\vspace{0.4em}
\textbf{(b) Version 2}
\end{minipage}

\caption{Conceptual workflows of the RAG systems evaluated in this work.}
\label{fig:workflow_vergleich}
\end{figure}

Beyond the core architecture, the system’s knowledge base was substantially expanded through an improved ingestion pipeline, increasing both coverage and metadata richness. V1 showed signigicant limitations in extracting course information which was adressed with this implementation. In addition, several complementary improvements were implemented to enhance overall system quality and user experience. First, response streaming was introduced, allowing generated tokens to be delivered incrementally to the user, thereby reducing perceived latency and improving interaction fluidity. Second, the system prompts were systematically redesigned following a structured prompt engineering framework \autocite{Lee2025PromptEngineering}, enabling clearer role definitions, improved context control, and more consistent output quality across different interaction modes. Finally, the citation mechanism was revised to increase transparency and usability by replacing abstract reference markers with human-readable source titles, allowing users to more easily assess the relevance and origin of retrieved information.