\section{Evaluation Framework}

\subsection{Data Basis}

\subsection{Multidimensional Evaluation Framework}

\paragraph{RAGAS: Automatic Evaluation Along the RAG Dimensions}
%references???
To enable a scalable and systematic assessment of the developed RAG chatbot, the evaluation was conducted using the Retrieval-Augmented Generation Assessment System (RAGAS) \cite{es2025ragas}. RAGAS was selected because it is a widely used framework for the automatic evaluation of RAG pipelines and allows quality to be assessed along the central dimensions of retrieval effectiveness, generation quality, and factual correctness. The framework is particularly suitable for applied settings, where large numbers of heterogeneous queries must be evaluated and manually created reference answers are not available.

Evaluating RAG systems differs from the evaluation of purely generative models, as the final output depends on both the retrieved context and the language model’s generation. Errors may originate from different stages of the pipeline, for example from retrieving irrelevant documents, from insufficient integration of multiple context passages, or from hallucinated statements despite the availability of correct evidence. Consequently, evaluation must consider both retrieval and generation performance, as well as the factual grounding of the produced answers.

RAGAS operationalizes this multi-stage perspective by providing automated metrics that quantify the effectiveness of the individual RAG components \cite{es2025ragas}. In this study, the metrics \textit{Answer Relevancy}, \textit{Context Relevance}, and \textit{Faithfulness} were used. \textit{Answer Relevancy} measures the semantic alignment between the user query and the generated response. \textit{Context Relevance} evaluates whether the retrieved context segments are suitable for answering the query and therefore reflects the effectiveness of the retrieval step. \textit{Faithfulness} assesses whether the generated answer is supported by the provided context and does not contain unsupported or hallucinated information. Together, these metrics allow a quantitative approximation of retrieval quality, generation quality, and factual correctness.

A key advantage of RAGAS is that it enables evaluation without predefined ground-truth answers. Instead of comparing system outputs to manually created reference responses, the framework uses a large language model as an evaluation instance (“LLM-as-a-judge”). This makes the evaluation scalable and allows real system responses to be assessed under realistic conditions. Such an approach is particularly relevant in educational environments, where large-scale manual annotation is difficult and only limited opportunities exist to perform controlled evaluations with reference answers.

The evaluation was based on a curated set of 60 questions representing the previously defined use-case categories, including technical support, course-related knowledge questions, and learning-support interactions. For each evaluation instance, the user query, the generated answer, and the context segments actually used during answer generation were passed to the RAGAS framework. This ensured that the evaluation reflected the real system behavior instead of an idealized retrieval output and allowed a precise comparison between different system versions.

To ensure stable results despite the stochastic nature of LLM-based evaluation, each question was processed multiple times and the resulting metric values were aggregated. The evaluation itself was performed using a separate large language model acting as the evaluation model, which enabled consistent scoring across all samples. The aggregated metric values were stored in structured form and used to compare the baseline system with the extended architecture.

While the automated RAGAS evaluation provides a quantitative view of retrieval effectiveness, answer relevance, and factual grounding, it cannot fully capture pedagogical quality or perceived usefulness, which are particularly relevant in educational applications. For this reason, RAGAS was used as the core component of the automatic evaluation and complemented by additional evaluation methods in the overall study design.