\section{Discussion}
% Präzisere Discussion --> hier kürzen
% Mehrwert des Ansatzes besser darstellen
% ERst Ergebnisse im Bildungskontext feststellen --> das dann auf eine RAG Ebene unabhängig von Kontext
% 3/4 Evaluation; 1/4 Lessons Learned für RAG-Systeme
\subsection{Reflection of the Evaluation Approach}

This study combines four evaluation methods: automated evaluation (RAGAS), expert and user evaluation as well as cost analysis to assess the quality of an RAG-based chatbot in an educational context as comprehensively as possible. This multidimensional comparative evaluation approach aligns recent research emphasizing that RAG evaluation must move beyond isolated component metrics toward system-level and application-oriented assessment \autocite{Yu2024RAGEvalSurvey,bdcc9120320}. This is particularly necessary in the field of education, as didactic and user-centered quality dimensions play a central role here alongside purely technical metrics. The results clearly show that each method has specific strengths and limitations, and only when used in combination do they provide a robust overall picture.

Automated evaluation using RAGAS offers advantages primarily in terms of scalability and comparability. Through the standardized calculation of metrics such as Answer Relevancy, Context Relevance and Faithfulness, large volumes of interactions can be efficiently evaluated, which reflects current trends toward LLM-based evaluation approaches \autocite{es2025ragas,saadfalcon2024ares,CRUDRAG}. At the same time, it becomes clear that these metrics are only of limited significance in an educational context. A central problem lies in the fact that RAGAS is primarily based on semantic similarities and thus primarily evaluates classic question-answer scenarios \autocite{Lewis2020_RAG}. Educationally relevant aspects such as didactic structure, comprehensibility, activation of prior knowledge, or the promotion of reflection, on the other hand, are not adequately captured. This limitation is further exacerbated by the specific use case. In learning environments, a response is not considered high-quality merely because it is factually correct and relevant, but because it supports learning processes. For example, intentionally simplified or abstracted responses can be pedagogically sound, even though they may appear less precise from a purely semantic evaluation perspective. Similarly, additional contextual information or explanatory elements that promote understanding may be incorrectly rated negatively by automated metrics. This demonstrates that for a robust evaluation in an educational context, either domain-specific, finely tuned evaluation models are required, or alternative approaches such as ground-truth-based evaluations must be used, even if these involve greater effort. Furthermore, complex forms of interaction, particularly the Socratic learning assistant developed in this work, cannot be adequately evaluated using this approach. However, it is precisely this form of dialogic, step-by-step knowledge transfer that represents a key added value of AI systems in an educational context. The sole use of automated metrics would therefore lead to a systematic underestimation of such functions.

Expert evaluation addresses precisely this gap and serves as an essential complement for a hybrid evaluation \autocite{bdcc9120320, 202504.0369}. By involving experts with backgrounds in both pedagogy and the subject matter, even complex interactions and qualitative aspects of system use can be assessed. In particular, the combination of standardized evaluations and think-aloud protocols enables a deeper understanding of perceived system quality, aligning with research emphasizing the importance of user-centered and context-aware evaluation in educational AI systems \autocite{ZawackiRichter2019}. The experts are able to evaluate answers not only in terms of their factual accuracy but also with regard to their didactic suitability and their contribution to the learning process. This is of central importance in an educational context, as the focus here is not only on the transmission of information but also on learning support. At the same time, it must be noted that this method involves a significant amount of effort and is subject to potential biases due to expert bias.

The user blind assessment in the form of a Chatbot Arena effectively supplements the evaluation approach by capturing user preference signals under controlled conditions.
While automated metrics like RAGAS struggle with qualitative dimensions, human evaluators naturally assess aspects such as tone, structural clarity, and overall comprehensibility.
However, interpreting these preferences in an educational context requires caution.
The arena measures perceived response quality rather than actual learning outcomes.
Evaluators may favor responses that are confidently formulated or structurally verbose, a cognitive bias in human preference testing of large language models documented, for example, by \autocite{Zheng2023LMSYSArena}.
Furthermore, users often prefer direct solutions, even though a less direct, Socratic approach where the system acts as a scaffold rather than an answer key has been shown to offer significantly higher pedagogical value \autocite{KASNECI2023102274}.
Additionally, the substantial rate of tied or rejected outcomes (17.5\% combined) indicates that technical architectural differences between the system variants (V1 and V2) often remain imperceptible to end users in practical application.

Like the RAGAS evaluation, the methodological design of the Arena presents specific limitations.
Relying on a predetermined set of single-turn questions ensures internal validity and allows for standardized A/B comparisons, but it limits external validity.
Restricting users from freely prompting the system prevents the evaluation of the model under fully realistic, dynamic conditions.
Most importantly, the single-turn, context-free design precludes the assessment of continuous dialogue and conversational memory.
As \autocite{Deriu2021} note, static single-turn evaluations are often inadequate for capturing the performance of multi-turn conversational systems.
Complex pedagogical interactions, such as those intended by the Socratic learning assistant developed in this work, cannot be adequately evaluated through isolated Q\&A pairs.
While the structured Arena provides a necessary baseline for general user acceptance, future evaluations of educational chatbots require scenario-based testing approaches.
Allowing users to engage in multi-turn, goal-oriented dialogues would provide a more accurate assessment of scaffolded knowledge transfer while maintaining measurable experimental control and dialogic knowledge transfer.

The cost analysis complements the qualitative and quantitative approaches by adding an economic perspective, which is becoming increasingly relevant for the practical deployment of educational technologies. Recent literature highlights that system-level design choices—such as retrieval strategies or reranking—directly impact computational efficiency and latency \autocite{liBPRAG,zhaoragSurvey}. Especially with scalable learning platforms, small differences in token consumption or latency can have a significant impact on operating costs and the user experience. However, this analysis should also be understood as indicative in its current form. A fully realistic assessment would need to account for additional factors, such as varying usage patterns, amortized infrastructure costs, or optimization strategies like prompt caching. Furthermore, it becomes apparent that aggregated metrics inadequately capture certain effects—such as savings from the workflow without retrieval.

Overall, the methodological reflection makes it clear that the evaluation of RAG systems in an educational context poses special challenges that go beyond traditional information retrieval and NLP metrics. An isolated examination of individual methods inevitably leads to oversimplified conclusions. Only through the combination of scalable automated procedures, domain-sensitive qualitative assessments, and economic analyses can a nuanced understanding of system quality be achieved. This insight underscores the need to align evaluation approaches more closely with the specific requirements of learning environments and evaluate using real use cases.

\subsection{Reflection of the Best Practice Implementation}

The results of this study demonstrate that the implemented architectural changes have varying effects on system performance and depend heavily on the specific usage context. The fundamental trade-off between improved quality and rising costs \autocite{liBPRAG,Neumann2024MoodleBot} becomes apparent.

A key influencing factor is the expansion of the knowledge base through the improved ingestion pipeline. While the increased information density leads to improved contextual relevance, the question arises as to whether the sharp rise in tokens actually results in a proportional gain in information or rather generates additional noise. Particularly in an educational context, where content is often already structured and didactically condensed, a further increase in the volume of data is not necessarily associated with better answer quality. This suggests that not only the quantity but especially the structure and didactic preparation of the data are decisive for the performance of RAG systems.

The introduction of the routing mechanism, on the other hand, represents one of the key positive developments. Through the differentiated processing of queries, the workflow without retrieval can be used efficiently to answer simple or conversational queries without additional retrieval effort. This adaptive approach is already reported as beneficial \autocite{liBPRAG,MultiHopRAG} and it not only reduces costs as well as latency but also improves the quality of responses by avoiding unnecessary context. However, the significance of this approach depends heavily on the usage scenario. In environments with a high proportion of simple interactions, this effect can be particularly pronounced.

With regard to established best practices from the RAG literature (hybrid retrieval, reranking, and multi-hop strategies) the results initially confirm their fundamental utility \autocite{liBPRAG,zhaoragSurvey}. The improvements in contextual relevance show that these components do indeed increase retrieval quality and thus enhance the system’s technical performance. At the same time, however, it becomes clear that these improvements do not necessarily lead to a corresponding perceived increase in quality in an educational context.

Particularly in traditional Q\&A scenarios, it becomes apparent that these approaches can sometimes be interpreted as overengineering. In learning management systems such as KI-Campus, information is often already available in content-rich, well-structured documents, meaning that relevant content can usually be found without complex multi-hop processing. In such cases, a strong contextual foundation achieved through additional retrieval steps is not strictly necessary. Rather, users sometimes seem to prefer more abstract, more summary-style answers that are not strictly tied to individual sources. This puts into perspective the blanket applicability of modern RAG best practices to the educational context and underscores the need for context-sensitive architectural decisions.

A more nuanced picture emerges with the Socratic learning assistant. Here, interactions clearly benefit from stronger contextual grounding and expanded retrieval strategies. By accessing a more comprehensive and precise knowledge base, the system can better understand what content is available to the learner and, building on that, ask targeted, guiding questions which benefit the user’s learning process \autocite{socaraticChatbot,Shetye2024Khanmigo}. This enables more coherent and highly individualized learning support while simultaneously reducing the risk of redundant or inconsistent explanations. In this context, the use of more complex, agent-based architectures also appears justified, as the additional flexibility offers clear pedagogical value. The results therefore suggest that the best-practice-inspired redesign particularly strengthens the chatbot’s role as a learning assistant, while trade-offs remain in technical support performance, language consistency, and perceived response speed.

In addition to structural changes, the results show that even relatively minor adjustments can have a significant impact on perceived system quality. In particular, improvements in citation functionality and the introduction of streaming were viewed positively by experts and contribute significantly to the user experience. This observation underscores that not only far-reaching architectural changes but also targeted UX optimizations are crucial to a system’s success. This aligns with broader research emphasizing the importance of usability and transparency in AI systems \autocite{ZawackiRichter2019}.

Overall, the results illustrate that modern RAG architectures must be viewed in a nuanced manner within an educational context. While best practices from the literature provide technically sound improvements, their actual added value depends heavily on the application scenario. A blanket adoption of these approaches is therefore not effective. Instead, the development of effective systems requires a conscious balancing of complexity, costs, and pedagogical benefits, as well as a clear alignment with the specific requirements of learning environments.