\section{Conclusion}
This study examines the development and comparative evaluation of a RAG-based chatbot in an educational context and demonstrates that architectural adjustments have varying effects on system quality and user perception. Based on the developed multidimensional comparative evaluation framework, it was demonstrated that improvements in ingestion and retrieval lead to significantly better contextual relevance, while the quality of the generated responses could not be consistently improved. The results rather illustrate that quality in an educational context should not be understood as monolithic, but rather depends heavily on the specific application scenario. In particular, the Socratic learning assistant proves to be a key added value of the enhanced architecture, as it enables didactically valuable interactions that go beyond classic Q\&A patterns. At the same time, the tension between quality gains and increased costs as well as latencies becomes apparent, underscoring the necessity of context-specific architectural decisions. However, the results support a shift away from rigid, linear workflows toward adaptive or even modular structures.
A key finding concerns the evaluation itself: No single method can comprehensively capture the quality of RAG systems in an educational context. Automated metrics are scalable but capture pedagogical quality only to a limited extent; expert evaluations provide nuanced insights but are time-consuming and potentially biased; user-based preference measurements reflect perceived quality, not necessarily quality that is effective for learning; finally, the cost analysis reveals that gains in quality must always be weighed against additional token consumption and higher latency. This leads to the key insight that a consistently hybrid evaluation is necessary, one that brings together technical, pedagogical, user-centered, and economic dimensions. At the same time, it becomes clear that the choice of evaluation methodology significantly shapes the results and is thus itself part of the system evaluation. Furthermore, realistic, use-case-based datasets are crucial for reliable conclusions, whereas benchmark-based approaches inadequately capture key usage scenarios.

Several avenues for future research emerge. First, the further development of dialogic, learning-oriented forms of interaction—particularly Socratic approaches—appears promising, as these have so far only been evaluable through limited automation. Second, there is a need for domain-specific evaluation frameworks that systematically assess pedagogical quality and go beyond existing metrics. Third, research on information density in vector databases in an educational context should be conducted to ensure optimal retrieval.

The limitations of this study lie, in particular, in its limited generalizability due to the specific application context (AI Campus), the limited number of system variants examined, and methodological constraints inherent in individual evaluation approaches. The evaluation is based predominantly on single-turn interactions, which means that complex dialogic learning processes are only partially captured. Despite these limitations, the study provides well-founded empirical and methodological contributions to the development and evaluation of RAG-based educational systems.